Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-11-17 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Update README

```{r compare_rain_ex_counties, echo = FALSE, fig.width = 10, fig.height = 10, efig.align = "center", fig.cap = "Comparison of rain estimates for example counties."}
plot_county_rain_compare <- function(ex_fips, ex_dir, ex_title, 
                                     get_data = FALSE){
        if(get_data){
                daily_timeseries(ex_fips, coverage = 0,
                                  date_min = "1988-01-01",
                                  date_max = "2011-12-31",
                                  var = "PRCP",
                                  out_directory = ex_dir)
        }

@@

        ave_n <- round(mean(ex_rain$ave_n, na.rm = TRUE))
        n_storms <- nrow(ex_rain)
        ex_title <- paste0(ex_title, "\n(storms: ", n_storms,
                           "; monitors:", ave_n, ")")

@@

a <- plot_county_rain_compare(ex_fips = "12086", ex_dir = "dade_data/",
                         ex_title = "Miami-Dade, FL")

b <- plot_county_rain_compare(ex_fips = "22071", ex_dir = "orleans_data/",
                         ex_title = "Orleans Parish, LA")

c <- plot_county_rain_compare(ex_fips = "45019", ex_dir = "charleston_data/",
                         ex_title = "Charleston County, SC")

d <- plot_county_rain_compare(ex_fips = "37183", ex_dir = "wake_data/",
                         ex_title = "Wake County, NC")

e <- plot_county_rain_compare(ex_fips = "13121", ex_dir = "fulton_data/",
                         ex_title = "Fulton County, GA")

f <- plot_county_rain_compare(ex_fips = "24005", ex_dir = "baltimore_data/",
                         ex_title = "Baltimore County, MD")

g <- plot_county_rain_compare(ex_fips = "42101",
                              ex_dir = "philadelphia_data/",
                         ex_title = "Philadelphia County, PA")

h <- plot_county_rain_compare(ex_fips = "48201", ex_dir = "harris_data/",
                         ex_title = "Harris County, TX")

i <- plot_county_rain_compare(ex_fips = "01097", ex_dir = "mobile_data/",
                         ex_title = "Mobile County, AL")

@@

```{r echo = FALSE, eval = FALSE}
options("noaakey" = Sys.getenv("noaakey"))

allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv") 
allison_fips <- sprintf("%05s", as.character(unique(allison_gauge$fips)))
# allison_fips <- sample(allison_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Allison-2001") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(allison_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/allison/")

load("MWR_gauge_rain_data/frances_gauge.RData")
frances_fips <- sprintf("%05s", as.character(unique(frances_gauge$fips)))
# frances_fips <- sample(frances_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Frances-2004") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(frances_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/frances/")

load("MWR_gauge_rain_data/ike_gauge.RData")
ike_fips <- sprintf("%05s", as.character(unique(ike_gauge$fips)))
# ike_fips <- sample(ike_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Ike-2008") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(ike_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/ike/")

load("MWR_gauge_rain_data/alex_gauge.RData")
alex_fips <- sprintf("%05s", as.character(unique(alex_gauge$fips)))
# alex_fips <- sample(alex_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Alex-2010") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(alex_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/alex/")
```

```{r county_ave_comparison, echo = FALSE, fig.width = 10, fig.height = 10, fig.align = "center", fig.cap = "Comparison of county-level average cumulative rainfall for the period from two days before to one day after the storm's closest approach to a county, using NLDAS-2 versus ground-based monitors, for counties with both estimates available for four example storms. The size of each point indicates the average number of monitors reporting from that county during the window (larger size indicates more monitors). The text over each point gives the FIPS county identifier."}
plot_county_ave_compare <- function(county_dir, storm_id){
        storm_title <- paste0(paste(unlist(strsplit(
                as.character(storm_id), "-")),
                             collapse = " ("), ")")
        storm_year <- gsub(".+-", "", storm_id)
        counties <- gsub(".rds", "",
                         list.files(paste0("county_ave/", county_dir)))

@@

        for(x in counties){
                county_weather <- readRDS(paste0("county_ave/", 
                                                 county_dir, "/", x,
                                                 ".rds"))$averaged %>%
                        dplyr::ungroup() %>%
                        dplyr::mutate(date = ymd(date)) %>%
                        dplyr::mutate(fips = x)
                storm_ave[[x]] <- county_weather
                }
        storm_ave <- do.call("rbind", storm_ave)

@@

        county_ave <- storm_ave %>%
                dplyr::left_join(storm_rain, by = "fips") %>%
                dplyr::group_by(fips) %>%
                dplyr::filter(ymd(closest_date) - ddays(2) <= date &
                       date <= ymd(closest_date) + ddays(1)) %>%
                dplyr::summarize(monitor_rain = sum(prcp),
                  tot_precip = first(tot_precip),
                  prcp_reporting = mean(prcp_reporting))
        plot_range <- range(county_ave[ , c("monitor_rain", "tot_precip")]) + 
                c(-15, 15)

@@

>>>>

```{r compare_rain_ex_counties, echo = FALSE, fig.width = 10, fig.height = 10, efig.align = "center", fig.cap = "Comparison of rain estimates for example counties."}
plot_county_rain_compare <- function(ex_fips, ex_dir, ex_title, 
                                     get_data = FALSE){
        if(get_data){
                daily_timeseries(ex_fips, coverage = 0,
                                  date_min = "1988-01-01",
                                  date_max = "2011-12-31",
                                  var = "PRCP",
                                  out_directory = ex_dir,
                                  keep_map = FALSE)
        }

@@

        ave_n <- round(mean(ex_rain$ave_n, na.rm = TRUE))
        n_storms <- nrow(ex_rain)
        ex_title <- paste0(ex_title, "\n(storms: ", n_storms,
                           "; monitors: ", ave_n, ")")

@@

a <- plot_county_rain_compare(ex_fips = "12086", ex_dir = "dade_data/",
                         ex_title = "Miami-Dade, FL",
                         get_data = FALSE)

b <- plot_county_rain_compare(ex_fips = "22071", ex_dir = "orleans_data/",
                         ex_title = "Orleans Parish, LA",
                         get_data = FALSE)

c <- plot_county_rain_compare(ex_fips = "45019", ex_dir = "charleston_data/",
                         ex_title = "Charleston County, SC",
                         get_data = FALSE)

d <- plot_county_rain_compare(ex_fips = "37183", ex_dir = "wake_data/",
                         ex_title = "Wake County, NC",
                         get_data = FALSE)

e <- plot_county_rain_compare(ex_fips = "13121", ex_dir = "fulton_data/",
                         ex_title = "Fulton County, GA",
                         get_data = FALSE)

f <- plot_county_rain_compare(ex_fips = "24005", ex_dir = "baltimore_data/",
                         ex_title = "Baltimore County, MD",
                         get_data = FALSE)

g <- plot_county_rain_compare(ex_fips = "42101",
                              ex_dir = "philadelphia_data/",
                         ex_title = "Philadelphia County, PA",
                         get_data = FALSE)

h <- plot_county_rain_compare(ex_fips = "48201", ex_dir = "harris_data/",
                         ex_title = "Harris County, TX",
                         get_data = FALSE)

i <- plot_county_rain_compare(ex_fips = "01097", ex_dir = "mobile_data/",
                         ex_title = "Mobile County, AL",
                         get_data = FALSE)

@@

```{r echo = FALSE, eval = FALSE}
allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv") 
allison_fips <- sprintf("%05s", as.character(unique(allison_gauge$fips)))
# allison_fips <- sample(allison_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Allison-2001") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
daily_timeseries(allison_fips, coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/allison/",
                 keep_map = FALSE)

load("MWR_gauge_rain_data/frances_gauge.RData")
frances_fips <- sprintf("%05s", as.character(unique(frances_gauge$fips)))
# frances_fips <- sample(frances_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Frances-2004") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
daily_timeseries(frances_fips, coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/frances/",
                 keep_map = FALSE)

load("MWR_gauge_rain_data/ike_gauge.RData")
ike_fips <- sprintf("%05s", as.character(unique(ike_gauge$fips)))
# ike_fips <- sample(ike_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Ike-2008") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
daily_timeseries(ike_fips, coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/ike/",
                 keep_map = FALSE)

load("MWR_gauge_rain_data/alex_gauge.RData")
alex_fips <- sprintf("%05s", as.character(unique(alex_gauge$fips)))
# alex_fips <- sample(alex_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Alex-2010") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
daily_timeseries(alex_fips, coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/alex/",
                 keep_map = FALSE)
```

```{r county_ave_comparison, echo = FALSE, fig.width = 10, fig.height = 10, fig.align = "center", fig.cap = "Comparison of county-level average cumulative rainfall for the period from two days before to one day after the storm's closest approach to a county, using NLDAS-2 versus ground-based monitors, for counties with both estimates available for four example storms. The size of each point indicates the average number of monitors reporting from that county during the window (larger size indicates more monitors). The text over each point gives the FIPS county identifier."}
plot_county_ave_compare <- function(county_dir, storm_id){
        storm_title <- paste0(paste(unlist(strsplit(
                as.character(storm_id), "-")),
                             collapse = " ("), ")")
        storm_year <- gsub(".+-", "", storm_id)
        counties <- gsub(".rds", "",
                         list.files(paste0("county_ave/", county_dir)))
        if(county_dir == "frances"){
                counties <- counties[counties != "51019"] # Missing all data
        }

@@

        for(x in counties){
                county_weather <- readRDS(paste0("county_ave/", 
                                                 county_dir, "/", x,
                                                 ".rds"))$daily_data %>%
                        dplyr::mutate(date = ymd(date),
                                      prcp = prcp / 10) %>%
                        dplyr::mutate(fips = x)
                storm_ave[[x]] <- county_weather
                }
        storm_ave <- do.call("rbind", storm_ave)

@@

        county_ave <- storm_ave %>%
                dplyr::left_join(storm_rain, by = "fips") %>%
                dplyr::group_by(fips) %>%
                dplyr::filter(ymd(closest_date) - ddays(2) <= date &
                       date <= ymd(closest_date) + ddays(1)) %>%
                dplyr::summarize(monitor_rain = sum(prcp),
                  tot_precip = first(tot_precip),
                  prcp_reporting = mean(prcp_reporting))
        plot_range <- range(county_ave[ , c("monitor_rain", "tot_precip")], 
                            na.rm = TRUE) + 
                c(-15, 15)

@@

```{r pareto_1, echo = FALSE, fig.width = 10, fig.height = 8, fig.align = "center", fig.cap = "Number of total county exposures summed across percentile bins of storms"}
my_pareto_1 <- function(data, title = "Rain exposure"){
        data <- data %>%
                select(storm_id, fips) %>%
                group_by(storm_id) %>%
                summarize(n = n()) %>%
                mutate(cat = ntile(n, 10)) %>%
                group_by(cat) %>%
                summarize(n = sum(n)) %>%
                ungroup() %>%
                arrange(desc(n)) %>%
                mutate(cumulative = cumsum(n),
                       group = 1:n(),
                       cat_labs = c("90-100th", "80-90th", "70-80th",
                               "60-70th", "50-60th", "40-50th",
                               "30-40th", "20-30th", "10-20th",
                               "0-10th")) %>%
                mutate(percent = paste0(round(100 * cumulative / sum(n)), "%"))
        my_pareto_plot <- ggplot(data, aes(x = group, y = n)) + 
                geom_bar(stat = "identity") + 
                geom_point(aes(x = group, y = cumulative), alpha = 0.25) + 
                geom_line(aes(x = group, y = cumulative), alpha = 0.25) + 
                geom_text(aes(y = cumulative, label = percent), 
                        hjust = "inward", vjust = "inward") + 
                theme_few() + 
                ylab("Number of\nexposed counties") + 
                ggtitle(title) + 
                scale_x_continuous(name = "", breaks = 1:10,
                                labels = data$cat_labs, 
                                trans = "reverse") +
                coord_flip()
        return(my_pareto_plot)
}

data <- county_distance(my_fips, start_year = 1988, end_year = 2011,
                               dist_limit = 100) 
a <- my_pareto_1(data, title = "Distance")

data <- county_rain(my_fips, start_year = 1988, end_year = 2011,
                               rain_limit = 75, dist_limit = 500) 
b <- my_pareto_1(data, title = "Rain")

data <- county_wind(my_fips, start_year = 1988, end_year = 2011,
                               wind_limit = 15) 
c <- my_pareto_1(data, title = "Wind")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "flood") 
d <- my_pareto_1(data, title = "Floods")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "tornado") 
e <- my_pareto_1(data, title = "Tornadoes")

grid.arrange(a, b, c, d, e, nrow = 2)
```

```{r pareto_2, echo = FALSE, fig.width = 10, fig.height = 8, fig.align = "center", fig.cap = "Number of total storm hits summed across percentile bins of counties"}
my_pareto_2 <- function(data, title = "Rain exposure"){
        data <- data %>%
                select(storm_id, fips) %>%
                group_by(fips) %>%
                summarize(n = n()) %>%
                mutate(cat = ntile(n, 10)) %>%
                group_by(cat) %>%
                summarize(n = sum(n)) %>%
                ungroup() %>%
                arrange(desc(n)) %>%
                mutate(cumulative = cumsum(n),
                       group = 1:n(),
                       cat_labs = c("90-100th", "80-90th", "70-80th",
                               "60-70th", "50-60th", "40-50th",
                               "30-40th", "20-30th", "10-20th",
                               "0-10th")) %>%
                mutate(percent = paste0(round(100 * cumulative / sum(n)), "%"))
        my_pareto_plot <- ggplot(data, aes(x = group, y = n)) + 
                geom_bar(stat = "identity") + 
                geom_point(aes(x = group, y = cumulative), alpha = 0.25) + 
                geom_line(aes(x = group, y = cumulative), alpha = 0.25) + 
                geom_text(aes(y = cumulative, label = percent), 
                        hjust = "inward", vjust = "inward") + 
                theme_few() + 
                ylab("Number of storm hits") + 
                ggtitle(title) + 
                scale_x_continuous(name = "", breaks = 1:10,
                                labels = data$cat_labs, 
                                trans = "reverse") +
                coord_flip()
        return(my_pareto_plot)
}

data <- county_distance(my_fips, start_year = 1988, end_year = 2011,
                               dist_limit = 100) 
a <- my_pareto_2(data, title = "Distance")

data <- county_rain(my_fips, start_year = 1988, end_year = 2011,
                               rain_limit = 75, dist_limit = 500) 
b <- my_pareto_2(data, title = "Rain")

data <- county_wind(my_fips, start_year = 1988, end_year = 2011,
                               wind_limit = 15) 
c <- my_pareto_2(data, title = "Wind")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "flood") 
d <- my_pareto_2(data, title = "Floods")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "tornado") 
e <- my_pareto_2(data, title = "Tornadoes")

grid.arrange(a, b, c, d, e, nrow = 2)
```

```{r power_1, echo = FALSE, fig.width = 10, fig.height = 8, fig.align = "center", fig.cap = "Distribution of number of exposed counties by rank of storm"}
my_power_1 <- function(data, title){
        data <- data %>%
                select(storm_id, fips) %>%
                group_by(storm_id) %>%
                summarize(n = n()) %>%
                ungroup() %>%
                arrange(desc(n)) %>%
                mutate(rank = 1:n())
        power_plot <- ggplot(data, aes(x = rank, y = n)) + 
                geom_point(alpha = 0.5) + 
                scale_y_continuous(trans = "log", breaks = c(1, 10, 100)) + 
                theme_few() + 
                ggtitle(title) + 
                xlab("Rank of storm") + ylab("# of counties exposed")
        return(power_plot)
}

data <- county_distance(my_fips, start_year = 1988, end_year = 2011,
                               dist_limit = 100) 
a <- my_power_1(data, title = "Distance")

data <- county_rain(my_fips, start_year = 1988, end_year = 2011,
                               rain_limit = 75, dist_limit = 500) 
b <- my_power_1(data, title = "Rain")

data <- county_wind(my_fips, start_year = 1988, end_year = 2011,
                               wind_limit = 15) 
c <- my_power_1(data, title = "Wind")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "flood") 
d <- my_power_1(data, title = "Floods")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "tornado") 
e <- my_power_1(data, title = "Tornadoes")

grid.arrange(a, b, c, d, e, nrow = 2)
```

```{r power_2, echo = FALSE, fig.width = 10, fig.height = 8, fig.align = "center", fig.cap = "Distribution of number of storm hits by rank of county"}
my_power_2 <- function(data, title){
        data <- data %>%
                select(storm_id, fips) %>%
                group_by(fips) %>%
                summarize(n = n()) %>%
                ungroup() %>%
                arrange(desc(n)) %>%
                mutate(rank = 1:n())
        power_plot <- ggplot(data, aes(x = rank, y = n)) + 
                geom_point(alpha = 0.5) + 
                scale_y_continuous(trans = "log", breaks = c(1, 10, 100)) + 
                theme_few() + 
                ggtitle(title) + 
                xlab("Rank of county") + ylab("# of storm hits")
        return(power_plot)
}

data <- county_distance(my_fips, start_year = 1988, end_year = 2011,
                               dist_limit = 100) 
a <- my_power_2(data, title = "Distance")

data <- county_rain(my_fips, start_year = 1988, end_year = 2011,
                               rain_limit = 75, dist_limit = 500) 
b <- my_power_2(data, title = "Rain")

data <- county_wind(my_fips, start_year = 1988, end_year = 2011,
                               wind_limit = 15) 
c <- my_power_2(data, title = "Wind")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "flood") 
d <- my_power_2(data, title = "Floods")

data <- county_events(my_fips, start_year = 1996, end_year = 2011,
                      event_type = "tornado") 
e <- my_power_2(data, title = "Tornadoes")

grid.arrange(a, b, c, d, e, nrow = 2)
```

```{r message = FALSE}
all_storm_counties <- closest_dist %>%
        dplyr::select(storm_id, fips)
metric_a <- county_rain(my_fips, start_year = 1988, end_year = 2011,
                               rain_limit = 75, dist_limit = 500) %>%
        select(storm_id, fips) %>%
        mutate(exposed_a = 1)
metric_b <- county_wind(my_fips, start_year = 1988, end_year = 2011,
                               wind_limit = 15) %>%
        select(storm_id, fips) %>%
        mutate(exposed_b = 1)
two_metrics <- metric_a %>%
        full_join(metric_b, by = c("storm_id", "fips")) %>%
        full_join(all_storm_counties, by = c("storm_id", "fips")) %>%
        mutate(exposed_a = ifelse(is.na(exposed_a), 0, 1),
               exposed_b = ifelse(is.na(exposed_b), 0, 1)) %>%
        group_by(storm_id) %>%
        summarize(a = sum(exposed_a & exposed_b),
                  b = sum(exposed_a & !exposed_b),
                  c = sum(!exposed_a & exposed_b),
                  d = sum(!exposed_a & !exposed_b),
                  f1 = sum(exposed_b),
                  f2 = sum(!exposed_b),
                  g1 = sum(exposed_a),
                  g2 = sum(!exposed_a),
                  n = n(),
                  no_exposure = a + b + c == 0,
                  prev = abs(a - d) / n,
                  bias = abs(b - c) / n,
                  prop_agree = (a + d) / n,
                  exp_agree = (f1 * g1 + f2 * g2) / n^2,
                  kappa = (prop_agree - exp_agree) / 
                                         (1 - exp_agree),
                  kappa = ifelse(is.nan(kappa) | is.infinite(kappa),
                                 0, kappa))

ggplot(two_metrics, aes(x = kappa, fill = no_exposure)) + 
        geom_histogram(color = "black")
```
